---
title: "Module 5: Conducting an ExWAS"
subtitle: "Conducting Exposome-Wide Association Studies"
author: "Chirag J Patel"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
library(tidyverse)
library(broom)
library(survey)
library(knitr)
library(kableExtra)
```

## Overview

This module covers the full ExWAS workflow:

1. ExWAS algorithm design (loop over exposures)
2. Manual association function (from Module 3)
3. Using `pe_flex_adjust()` at scale
4. Error handling with `tryCatch()`
5. Collecting results into a tibble
6. FDR correction
7. Train/test replication
8. Parallelization with `furrr`
9. Binary phenotype ExWAS with logistic regression

## The ExWAS Algorithm

For each exposure $E_j$ in the exposure catalog:

1. Retrieve phenotype and exposure data
2. Check exposure data type (continuous vs. categorical)
3. Fit survey-weighted regression: $P \sim E_j + \text{covariates}$
4. Extract coefficient, standard error, p-value
5. Collect into results table
6. Apply FDR correction across all exposures

## Recap: The Association Function

From Module 3, our manual association function (using bundled data):

```{r manual-association}
library(survey)
load('../data/nh.Rdata')

dsn <- svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                 weights = ~WTMEC2YR, nest = TRUE, data = NHData.train)

association <- function(phenotype = "BMXBMI", exposure = "LBXBCD",
                        covariates = c("RIDAGEYR", "male",
                                      "black", "mexican",
                                      "other_hispanic", "other_eth",
                                      "INDFMPIR"),
                        dsn) {
  covariate_string <- paste(covariates, collapse = " + ")
  mod_string <- sprintf("scale(%s) ~ scale(log(%s + 0.001)) + %s",
                        phenotype, exposure, covariate_string)
  mod <- svyglm(as.formula(mod_string), dsn)
  return(mod)
}
```

## Testing the Association Function

```{r test-association}
# Test: BMI ~ Blood Cadmium
test_model <- association("BMXBMI", "LBXBCD", dsn = dsn)
tidy(test_model) %>%
  kable(digits = 4) %>% kable_styling(font_size = 9)
```

## Building the ExWAS Loop (Manual Approach)

```{r exwas-loop-manual}
# ExWAS on BMI using the training data
results_train <- tibble()

for (exp_var in ExposureList) {
  result <- tryCatch({
    mod <- association("BMXBMI", exp_var, dsn = dsn)
    mod_tidy <- tidy(mod)
    mod_glance <- glance(mod)

    # Extract the exposure coefficient (2nd row: scale(log(...)))
    exp_row <- mod_tidy %>%
      filter(grepl("scale\\(log", term))

    tibble(
      exposure = exp_var,
      estimate = exp_row$estimate,
      std.error = exp_row$std.error,
      p.value = exp_row$p.value,
      r.squared = mod_glance$adj.r.squared,
      n = mod_glance$nobs
    )
  }, error = function(e) {
    tibble(
      exposure = exp_var,
      estimate = NA_real_,
      std.error = NA_real_,
      p.value = NA_real_,
      r.squared = NA_real_,
      n = NA_integer_
    )
  })

  results_train <- bind_rows(results_train, result)
}
```

## Why tryCatch()?

Some exposures may fail for various reasons:

- Too few non-missing observations
- Convergence issues in the model
- Zero variance in the exposure

`tryCatch()` prevents the entire loop from crashing:

```{r trycatch-demo, eval=FALSE}
tryCatch({
  # Code that might fail
  mod <- svyglm(formula, dsn)
  tidy(mod)
}, error = function(e) {
  # Return NA row on failure
  tibble(estimate = NA, p.value = NA)
})
```

## A Cleaner Loop with purrr::map

```{r exwas-purrr}
run_exwas <- function(exposure_list, phenotype, dsn) {
  map_dfr(exposure_list, function(exp_var) {
    tryCatch({
      mod <- association(phenotype, exp_var, dsn = dsn)
      mod_tidy <- tidy(mod) %>% filter(grepl("scale\\(log", term))
      mod_glance <- glance(mod)
      tibble(
        exposure = exp_var,
        estimate = mod_tidy$estimate,
        std.error = mod_tidy$std.error,
        p.value = mod_tidy$p.value,
        r.squared = mod_glance$adj.r.squared,
        n = mod_glance$nobs
      )
    }, error = function(e) {
      tibble(exposure = exp_var, estimate = NA_real_,
             std.error = NA_real_, p.value = NA_real_,
             r.squared = NA_real_, n = NA_integer_)
    })
  })
}
```

## Running ExWAS on BMI

```{r exwas-bmi}
# Training set
results_bmi_train <- run_exwas(ExposureList, "BMXBMI", dsn)

# Testing set
dsn_test <- svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                      weights = ~WTMEC2YR, nest = TRUE, data = NHData.test)
results_bmi_test <- run_exwas(ExposureList, "BMXBMI", dsn_test)
```

## Running ExWAS on Glucose

```{r exwas-glucose}
# Training set
results_glu_train <- run_exwas(ExposureList, "LBXGLU", dsn)

# Testing set
results_glu_test <- run_exwas(ExposureList, "LBXGLU", dsn_test)
```

## FDR Correction

Apply Benjamini-Hochberg FDR to training set p-values:

```{r fdr-correction}
results_bmi_train <- results_bmi_train %>%
  filter(!is.na(p.value)) %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

results_glu_train <- results_glu_train %>%
  filter(!is.na(p.value)) %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))
```

## Top Hits: BMI

```{r top-bmi}
results_bmi_train %>%
  arrange(fdr) %>%
  head(10) %>%
  left_join(ExposureDescription, by = c("exposure" = "var")) %>%
  select(exposure, var_desc, estimate, fdr) %>%
  kable(digits = 4) %>% kable_styling(font_size = 9)
```

## Top Hits: Glucose

```{r top-glucose}
results_glu_train %>%
  arrange(fdr) %>%
  head(10) %>%
  left_join(ExposureDescription, by = c("exposure" = "var")) %>%
  select(exposure, var_desc, estimate, fdr) %>%
  kable(digits = 4) %>% kable_styling(font_size = 9)
```

## Replication Strategy

**Train/test replication** validates findings in independent data:

1. **Discovery** (training set): FDR < 0.05
2. **Replication** (testing set): p-value < 0.05
3. **Concordance**: same direction of effect in both sets

This reduces false positives beyond FDR correction alone.

## Implementing Replication

```{r replication}
# Join training and testing results
replicated_bmi <- results_bmi_train %>%
  filter(fdr < 0.05) %>%
  select(exposure, est_train = estimate, fdr_train = fdr) %>%
  inner_join(
    results_bmi_test %>%
      select(exposure, est_test = estimate, p_test = p.value),
    by = "exposure"
  ) %>%
  filter(
    p_test < 0.05,                        # significant in test
    sign(est_train) == sign(est_test)      # concordant direction
  )
```

## Replication Results: BMI

```{r replication-results}
replicated_bmi %>%
  left_join(ExposureDescription, by = c("exposure" = "var")) %>%
  select(exposure, var_desc, est_train, est_test, fdr_train, p_test) %>%
  arrange(fdr_train) %>%
  kable(digits = 4) %>% kable_styling(font_size = 8)
```

## Using nhanespewas for ExWAS

Now let's do the same using the `nhanespewas` package with the full database:

```{r nhanespewas-exwas}
library(nhanespewas)
con <- connect_pewas_data()

# Get list of exposures from e_catalog
exposure_vars <- e_catalog$var
```

## ExWAS with pe_flex_adjust()

```{r pe-flex-loop}
run_exwas_pewas <- function(exposure_vars, phenotype, adj_model, con) {
  map_dfr(exposure_vars, function(exp_var) {
    tryCatch({
      result <- pe_flex_adjust(
        phenotype = phenotype,
        exposure = exp_var,
        adjustment_model = adj_model,
        con = con
      )
      # Extract results from all cycles
      result %>%
        map_dfr(~ tidy(.), .id = "cycle") %>%
        filter(grepl(exp_var, term)) %>%
        summarize(
          exposure = exp_var,
          estimate = mean(estimate),
          std.error = mean(std.error),
          p.value = min(p.value),
          n_cycles = n()
        )
    }, error = function(e) {
      tibble(exposure = exp_var, estimate = NA_real_,
             std.error = NA_real_, p.value = NA_real_,
             n_cycles = NA_integer_)
    })
  })
}
```

## Running ExWAS with nhanespewas

```{r run-pewas-exwas}
# Run ExWAS for BMI with adjustment model 4
results_pewas_bmi <- run_exwas_pewas(
  exposure_vars = exposure_vars,
  phenotype = "BMXBMI",
  adj_model = adjustment_models[[4]],
  con = con
)

# Apply FDR
results_pewas_bmi <- results_pewas_bmi %>%
  filter(!is.na(p.value)) %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))
```

## Parallelization with furrr

For large exposure lists, parallelize the loop:

```{r furrr-setup}
library(furrr)
library(future)

# Set up parallel workers
plan(multisession, workers = 4)
```

## Parallel ExWAS

```{r furrr-exwas}
run_exwas_parallel <- function(exposure_list, phenotype, dsn) {
  future_map_dfr(exposure_list, function(exp_var) {
    tryCatch({
      mod <- association(phenotype, exp_var, dsn = dsn)
      mod_tidy <- tidy(mod) %>% filter(grepl("scale\\(log", term))
      mod_glance <- glance(mod)
      tibble(
        exposure = exp_var,
        estimate = mod_tidy$estimate,
        std.error = mod_tidy$std.error,
        p.value = mod_tidy$p.value,
        r.squared = mod_glance$adj.r.squared,
        n = mod_glance$nobs
      )
    }, error = function(e) {
      tibble(exposure = exp_var, estimate = NA_real_,
             std.error = NA_real_, p.value = NA_real_,
             r.squared = NA_real_, n = NA_integer_)
    })
  }, .options = furrr_options(seed = TRUE))
}
```

## Running in Parallel

```{r run-parallel}
# Run ExWAS in parallel (4 cores)
results_parallel <- run_exwas_parallel(ExposureList, "BMXBMI", dsn)

# Reset to sequential
plan(sequential)
```

The parallel version should be ~3-4x faster with 4 workers.

## Binary Phenotype ExWAS: Logistic Regression

For binary outcomes (e.g., T2D status), use logistic regression:

```{r binary-outcome-setup}
# Create binary T2D variable (glucose >= 126)
NHData.train <- NHData.train %>%
  mutate(t2d = as.integer(LBXGLU >= 126))

# Rebuild design
dsn <- svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                 weights = ~WTMEC2YR, nest = TRUE, data = NHData.train)
```

## Logistic Association Function

```{r logistic-association}
logistic_association <- function(outcome = "t2d", exposure = "LBXBCD",
                                 covariates = c("RIDAGEYR", "RIAGENDR",
                                               "factor(RIDRETH1)", "INDFMPIR"),
                                 dsn) {
  covariate_string <- paste(covariates, collapse = " + ")
  mod_string <- sprintf("%s ~ scale(log(%s + 0.001)) + %s",
                        outcome, exposure, covariate_string)
  mod <- svyglm(as.formula(mod_string), dsn, family = quasibinomial())
  return(mod)
}
```

## Using nhanespewas for Logistic ExWAS

```{r logistic-pewas}
# nhanespewas provides logistic_e_flex_adjust()
result_logistic <- logistic_e_flex_adjust(
  phenotype = "t2d_status",  # binary phenotype variable
  exposure = "LBXBPB",
  adjustment_model = adjustment_models[[4]],
  con = con
)
```

## Interpreting Logistic Results

```{r logistic-interpret}
# Coefficients are log-odds ratios
# Exponentiate to get odds ratios
result_logistic %>%
  map_dfr(~ tidy(.), .id = "cycle") %>%
  filter(grepl("LBXBPB", term)) %>%
  mutate(odds_ratio = exp(estimate)) %>%
  select(cycle, estimate, odds_ratio, p.value) %>%
  kable(digits = 3) %>% kable_styling()
```

An OR > 1 means higher exposure is associated with higher odds of the outcome.

## ExWAS Results Summary

After running a full ExWAS, your results tibble should contain:

| Column | Description |
|--------|-------------|
| `exposure` | Exposure variable name |
| `estimate` | Coefficient (per 1 SD change in log-exposure) |
| `std.error` | Standard error of the estimate |
| `p.value` | Raw p-value |
| `fdr` | FDR-corrected p-value |
| `r.squared` | Model R² |
| `n` | Sample size |

## Counting Significant Findings

```{r count-significant}
results_bmi_train %>%
  summarize(
    total_tested = n(),
    bonferroni_sig = sum(p.value < 0.05 / n(), na.rm = TRUE),
    fdr_sig = sum(fdr < 0.05, na.rm = TRUE),
    nominal_sig = sum(p.value < 0.05, na.rm = TRUE)
  ) %>%
  kable() %>% kable_styling()
```

## Cleaning Up

```{r cleanup}
disconnect_pewas_data(con)
```

## Summary

- The ExWAS **loop** tests each exposure against a phenotype with proper covariates
- **`tryCatch()`** prevents individual failures from crashing the pipeline
- **FDR correction** controls false discoveries across all tests
- **Replication** in independent data adds a second layer of validation
- **`pe_flex_adjust()`** provides a streamlined interface through `nhanespewas`
- **Parallelization** with `furrr` speeds up computation
- **Logistic regression** extends ExWAS to binary outcomes

## What's Next?

**Module 6: Interpreting ExWAS Results**

- Volcano plots and effect size visualization
- Delta R-squared analysis
- Replication tables
- Exposure correlation structure
- Sensitivity analysis across adjustment models
